Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-06-25 writeup/pnas-word-order-si.tex
code

\textcolor{red}{TODOs}

\textcolor{red}{run czech parser}

\textcolor{red}{keep in mind: pareto plot unify with SI plot}

@@

The basis for our precise formalization is the function proposed in \cite{ferreri2003least,ferrericancho2007global,futrell2017memory,peloquin2019interactions} as a general efficiency metric for communicative systems.
%The key idea is to maximize the \key{amount of information} that linguistic forms provide about meanings, while constraining \key{complexity and diversity} of forms. %; there are some differences in the formalization of complexity.
If $S$ denotes signals (e.g., words, sentences) and $R$ denotes their referents (e.g., objects in a reference game), then this efficiency metric takes the form (notation slightly varies across these publications):
\begin{equation}\label{eq:eff-general}
I[S, R] - \lambda H[S]
\end{equation}
where $I[S, R]$ describes the \key{informativity} of the signals $S$ about their referents $R$, and $H[S]$ describes the \key{complexity} of the communication system.
While prior studies \cite{ferreri2003least, kemp2012kinship,regier2015word,zaslavsky2018efficient} mostly considered settings where the signals $S$ are individual words without further structure, the signals are entire sentences $\utterance$ in our setting.
The underlying messages $R$ which the speaker aims to convey are the syntactic structures $\tree$.
By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined.
The syntactic structure specifies how the meanings of words are combined; therefore, recovering the syntactic structure is a prerequisite to understanding a sentence correctly.
%The idea that efficient languages will permit the listener to infer grammati

We therefore arrive at the following efficiency metric:
\begin{equation}\label{eq:efficiency-derived}
	R_{\textit{Eff}} := R_{\textit{Parseability}} + \lambda R_\textit{Pred}
\end{equation}
where \key{parseability} is the amount of information that utterances provide about their underlying syntactic structures:
\begin{equation}
	R_{Pars} := I[\utterance,\tree] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
and \key{predictability} is the negative entropy or surprisal of the language:
\begin{equation}
	R_{Pred} := - H[\utterance] = \sum_{u} p(u) \log p(u)
\end{equation}

We now discuss how this metric relates to other variants that have been proposed in the information-theoretic literature on language, and describe why~(\ref{eq:eff-general}-\ref{eq:efficiency-derived}) is particularly appropriate to our setting.
This efficiency metric~(\ref{eq:eff-general}-\ref{eq:efficiency-derived}) is equivalent to a deterministic version \cite{strouse2017deterministic} of the Information Bottleneck approach, which has been successfully applied to modeling word meaning across different domains \cite{zaslavsky2018efficient, zaslavsky2019semantic}.
The Information Bottleneck models complexity using a mutual information term instead of the entropy appearing in~(\ref{eq:eff-general}).
This mutual information formalization is appropriate to codes that are nondeterministic \cite{zaslavsky2018efficient}.
As we assume deterministic grammars that transduce every underlying syntactic structure into one surface order, our model corresponds to the Deterministic Information Bottleneck~\cite{strouse2017deterministic}, which arises when the encoder is deterministic, and which uses the entropy as the complexity measure.
Entropy also corresponds to the average of surprisal, which is a strong and linear predictor of human language processing effort~\cite{hale2001probabilistic,levy2008expectation,smith2013effect}.

@@

@@

In the efficiency objective
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Pars}} + \lambda R_\textit{Pred}
\end{equation}
the value of $\lambda$ is constrained to be in $[0,1)$.
This means, surprisal must be weighted less strongly than parseability.

@@

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parseability.
The computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$.
However, we also examined word order predictions for grammars optimized for only parseability or only predictability, in order to tease apart predictions made by these two components.
As shown in Table~\ref{tab:all-predictions-1}, each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions.
That is, at $\lambda$ close to its maximal value, the predictions of optimizing the two scoring functions individually add up to the predictions of efficiency optimization.\footnote{Results from one of the preliminary experiments reported in Figure~\ref{table:corr-resu-previous} show that results are stable to small variation of $\lambda$: Essentially equivalent predictions are obtained for $\lambda=1.0$. While $\lambda=1.0$ is not a valid choice for communicative efficiency in general due to the possibility of collapse to a single utterance, our setting does not allow such a collapse, as the syntactic structure already determines which words are present in the sentence.}
Small values of $\lambda$ correspond to the case where predictability plays no role, and only parseability is optimized (Table~\ref{tab:all-predictions-1}), in which case not all correlations are predicted (Figure~\ref{fig:posterior}).
This is confirmed by converging evidence from our preregistered preliminary experiments in Figure~\ref{table:corr-resu-previous}.

@@

Figure 4 in the main paper shows the average of these per-language plots.

@@

\subsection{Previous Experiments}\label{sec:previous-exps}
In Table~\ref{table:corr-resu-previous} we report the results of our two previous, preregistered, simulations\footnote{\url{http://aspredicted.org/blind.php?x=8gp2bt}, \url{https://aspredicted.org/blind.php?x=bg35x7}. For the results of the locality simulations described in the first preregistration, see the Dependency Length Minimization results in Table~\ref{tab:all-predictions-1a}, with discussion in Section~\ref{sec:DLM}.} together with results from the main experiment.
These experiments all had the same setup described in Section~\ref{sec:neural-architectures}, which was fixed before starting simulations; differences are that (1) one simulation places fully equal weight on parseability and predictability ($\lambda=1.0$), and (2) the final experiment uses three random seeds per grammar.
Results across all three experiments agree; jointly optimizing grammars for parseability and predictability produces all eight correlations.

@@

\subsection{Extracting Grammars from Datasets}\label{sec:extract-grammars}
We extract grammars for the actual languages by fitting a differentiable ordering grammar maximizing the likelihood of the observed orderings.
To prevent overfitting, we regularize each $a_\tau$, $b_\tau$ with a simple Bayesian prior $logit(a_\tau) \sim \mathcal{N}(0,1)$, $b_\tau \sim \mathcal{N}(0,1)$.\footnote{We implemented this regularized optimization as mean-field ELBO variational inference in Pyro~\cite{bingham2018pyro}.}
We then extract the posterior means for each parameter $a_\tau, b_\tau$, and convert the resulting differentiable grammar into an ordinary ordering grammar.

>>>>

@@

The basis for our precise formalization is the function proposed in \cite{ferreri2003least,ferrericancho2007global,futrell2017memory,peloquin2019interactions} as a general efficiency metric for communicative systems.
%The key idea is to maximize the \key{amount of information} that linguistic forms provide about meanings, while constraining \key{complexity and diversity} of forms. %; there are some differences in the formalization of complexity.
If $S$ denotes signals (e.g., words, sentences) and $R$ denotes their referents (e.g., objects in a reference game), then this efficiency metric takes the form (notation slightly varies across these publications):
\begin{equation}\label{eq:eff-general}
I[S, R] - \lambda \cdot H[S]
\end{equation}
where $I[S, R]$ describes the \key{informativity} of the signals $S$ about their referents $R$, and $H[S]$ describes the \key{complexity} of the communication system, and $\lambda > 0$ trades off the two aspects of efficiency.
While prior studies \cite{ferreri2003least, kemp2012kinship,regier2015word,zaslavsky2018efficient} mostly considered settings where the signals $S$ are individual words without further structure, the signals are entire sentences $\utterance$ in our setting.
The underlying messages $R$ which the speaker aims to convey are the syntactic structures $\tree$.
By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined.
The syntactic structure specifies how the meanings of words are combined; therefore, recovering the syntactic structure is a prerequisite to understanding a sentence correctly.
%The idea that efficient languages will permit the listener to infer grammati
Hence, substituting utterances $\utterance$ for signals $S$, and syntactic structures $\tree$ for underlying messages $R$, into (\ref{eq:eff-general}), we  arrive at the following efficiency metric for word order:
\begin{equation}\label{eq:efficiency-derived}
	R_{\textit{Eff}} := R_{\textit{Pars}} + \lambda \cdot R_\textit{Pred}
\end{equation}
where \key{parseability} is the amount of information that utterances provide about their underlying syntactic structures:
\begin{equation}
	R_{Pars} := I[\utterance,\tree] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
\end{equation}
and \key{predictability} is the negative entropy or surprisal of the language:
\begin{equation}
	R_{Pred} := - H[\utterance] = \sum_{u} p(u) \log p(u)
\end{equation}

We now discuss how this metric relates to other variants that have been proposed in the information-theoretic literature on language, and describe why~(\ref{eq:eff-general}-\ref{eq:efficiency-derived}) is most appropriate to our setting.
This efficiency metric~(\ref{eq:eff-general}-\ref{eq:efficiency-derived}) is equivalent to a deterministic version \cite{strouse2017deterministic} of the Information Bottleneck approach, which has been successfully applied to modeling word meaning across different domains \cite{zaslavsky2018efficient, zaslavsky2019semantic}.
The Information Bottleneck models complexity using a mutual information term instead of the entropy appearing in~(\ref{eq:eff-general}).
This mutual information formalization is appropriate to codes that are nondeterministic \cite{zaslavsky2018efficient}.
As we assume deterministic grammars that transduce every underlying syntactic structure into one surface order, our model corresponds to the Deterministic Information Bottleneck~\cite{strouse2017deterministic}, which arises when the encoder is deterministic, and which uses the entropy as the complexity measure.
Entropy also corresponds to the average of surprisal, which is a strong and linear predictor of human language processing effort~\cite{hale2001probabilistic,levy2008expectation,smith2013effect}.

@@

Beyond the study of natural language, the eficiency metric (\ref{eq:eff-general}) is also closely related to information-theoretic models in other disciplines.
The tradeoff between informativity and complexity of communication systems is studied by Rate-distortion theory \cite{shannon1948mathematical, berger1971rate}.
The \emph{Infomax Principle} derives parsimonious data representations by maximizing the mutual information between data and representations, subject to constraints on the representations~\cite{linsker1988self}; a constraint on the representation entropy leads to a metric equivalent to (\ref{eq:eff-general}) \cite{friston2010free}.

@@

In the efficiency objective (\ref{eq:efficiency-derived})
\begin{equation}\label{eq:efficiency}
	R_{\textit{Eff}} := R_{\textit{Pars}} + \lambda R_\textit{Pred}
\end{equation}
the value of $\lambda$ is constrained to be in $[0,1)$.
This means, surprisal must be weighted less strongly than parseability.

@@

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parseability.
While the computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$, we also examined word order predictions for grammars optimized for only parseability or only predictability, in order to tease apart predictions made by these two components.
As shown in Table~\ref{tab:all-predictions-1}, each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions.
That is, at $\lambda$ close to its maximal value, the predictions of optimizing the two scoring functions individually add up to the predictions of efficiency optimization.\footnote{Results from one of the preliminary experiments reported in Figure~\ref{table:corr-resu-previous} show that results are stable to small variation of $\lambda$: Essentially equivalent predictions are obtained for $\lambda=1.0$. While $\lambda=1.0$ is not a valid choice for communicative efficiency in general due to the possibility of collapse to a single utterance, our setting does not allow such a collapse, as the syntactic structure already determines which words are present in the sentence.}
Small values of $\lambda$ correspond to the case where predictability plays no role, and only parseability is optimized (Table~\ref{tab:all-predictions-1}), in which case not all correlations are predicted (Figure~\ref{fig:posterior}).
This is confirmed by converging evidence from our preregistered preliminary experiments in Figure~\ref{table:corr-resu-previous}.

@@

Figure 4 in the main paper shows the average of these per-language plots, with a kernel density estimate of the distribution of baseline grammars.

@@

\subsection{Previous Experiments}\label{sec:previous-exps}
In Table~\ref{table:corr-resu-previous} we report the results of our two previous, preregistered, simulations\footnote{\url{http://aspredicted.org/blind.php?x=8gp2bt}, \url{https://aspredicted.org/blind.php?x=bg35x7}. For the results of the Locality simulations described in the first preregistration, see the Dependency Length Minimization results in Table~\ref{tab:all-predictions-1b}, with discussion in Section~\ref{sec:DLM}.} together with results from the main experiment.
These experiments all had the same setup described in Section~\ref{sec:neural-architectures}, which was fixed before starting simulations; differences are that (1) one simulation places fully equal weight on parseability and predictability ($\lambda=1.0$), and (2) the final experiment uses three random seeds per grammar.
Results across all three experiments agree; jointly optimizing grammars for parseability and predictability produces all eight correlations.

@@

\subsection{Extracting Grammars from Datasets}\label{sec:extract-grammars}
We extract grammars for the actual languages by fitting a differentiable ordering grammar maximizing the likelihood of the observed orderings.
To prevent overfitting, we regularize each $a_\tau$, $b_\tau$ with a simple Bayesian prior $logit(a_\tau) \sim \mathcal{N}(0,1)$, $b_\tau \sim \mathcal{N}(0,1)$.
We implemented this regularized optimization as mean-field ELBO variational inference in Pyro~\cite{bingham2018pyro}.
We then extract the posterior means for each parameter $a_\tau, b_\tau$, and convert the resulting differentiable grammar into an ordinary ordering grammar.
